# Results

This section presents the results of our classification models across combinations of crash types (all, bicycle-involved, pedestrian-involved), outcome variables (fatal, injury, major injury), predictor sets (all features, infrastructure-only), and modeling approaches (Logistic Regression, LASSO, Ridge, Random Forest). Our focus is on model interpretability, with an emphasis on feature importance rankings rather than predictive performance alone. Cohen’s Kappa is used as the primary evaluation metric due to class imbalance. For each modeling configuration, we report predictive agreement and the top three contributing features. A detailed table view is presented in the Appendix.

Some of the models have more than one feature per a particular rank in the top three. These are ties based on the most common occurrence among the five model runs that were completed. 

The LASSO model has the power to shrink some coefficients on predictors to zero. If this occurs for every predictor in the dataset, then no features will show up in the top three. This occurs in some of the bike/pedestrian data subsets.

```{r}
library(patchwork)
results_agg <- read_csv("data/modeling_results_agg.csv")
results_heatmap <- results_agg %>%
  filter(`Predictors Subset` == "All") %>%
  pivot_longer(cols = c(`Highest Importance`, `2nd Highest Importance`, `3rd Highest Importance`), names_to = "Rank", values_to = "Feature") %>%
  filter(Feature != "") %>%
  mutate(Feature = str_split(Feature, ",")) %>%
  unnest(Feature) %>%
  mutate(Feature = str_trim(Feature)) %>%
  mutate(
    Importance_Rank = case_when(
      Rank == "Highest Importance" ~ 1,
      Rank == "2nd Highest Importance" ~ 2,
      Rank == "3rd Highest Importance" ~ 3
    ),
    Model_ID = paste(`Model Type`, `Target Variable`, sep = " - ")
  ) %>%
  group_by(`Data Subset`, Model_ID, Feature) %>%
  slice_min(order_by = Importance_Rank, with_ties = TRUE) %>%
  ungroup()
results_heatmap_infra <- results_agg %>%
  filter(`Predictors Subset` == "Infrastructural") %>%
  pivot_longer(cols = c(`Highest Importance`, `2nd Highest Importance`, `3rd Highest Importance`), names_to = "Rank", values_to = "Feature") %>%
  filter(Feature != "") %>%
  mutate(Feature = str_split(Feature, ",")) %>%
  unnest(Feature) %>%
  mutate(Feature = str_trim(Feature)) %>%
  mutate(
    Importance_Rank = case_when(
      Rank == "Highest Importance" ~ 1,
      Rank == "2nd Highest Importance" ~ 2,
      Rank == "3rd Highest Importance" ~ 3
    ),
    Model_ID = paste(`Model Type`, `Target Variable`, sep = " - ")
  ) %>%
  group_by(`Data Subset`, Model_ID, Feature) %>%
  slice_min(order_by = Importance_Rank, with_ties = TRUE) %>%
  ungroup()
results_heatmap$Feature <- factor(results_heatmap$Feature, levels = c(
  # Driver Behavior
  "Alcohol Involved", "Drugs Involved", "Marijuana Involved", "Crash Speed Involved", "Hit and Run",
  # Environment
  "Road Surface: Snow", "Road Surface: Ice", "Road Surface: Wet", "Road Surface: Unknown", 
  "Road Character: Driveway or Alley", "Weather Condition: Snow",
  # Human Influenced Factors
  "Dark- No Street Lights", "Total Watts", "Speed Limit", "Sum Traffic Calming Devices", 
  # Other
  "Crash Hour", "Weekday"
))
results_heatmap$Model_ID <- factor(results_heatmap$Model_ID, levels = c(
  "Logistic - Injured vs Non-Injured", "LASSO - Injured vs Non-Injured", 
  "Ridge - Injured vs Non-Injured",  "Random Forest - Injured vs Non-Injured", 
  "Logistic - Degree of Injury", "LASSO - Degree of Injury", 
  "Ridge - Degree of Injury", "Random Forest - Degree of Injury",
  "Logistic - Fatal vs Non-Fatal", "LASSO - Fatal vs Non-Fatal", 
  "Ridge - Fatal vs Non-Fatal", "Random Forest - Fatal vs Non-Fatal"
))
results_heatmap_infra$Model_ID <- factor(results_heatmap_infra$Model_ID, levels = c(
  "Logistic - Injured vs Non-Injured", "LASSO - Injured vs Non-Injured", 
  "Ridge - Injured vs Non-Injured",  "Random Forest - Injured vs Non-Injured", 
  "Logistic - Degree of Injury", "LASSO - Degree of Injury", 
  "Ridge - Degree of Injury", "Random Forest - Degree of Injury",
  "Logistic - Fatal vs Non-Fatal", "LASSO - Fatal vs Non-Fatal", 
  "Ridge - Fatal vs Non-Fatal", "Random Forest - Fatal vs Non-Fatal"
))
```

## Modeling Results - All Crashes
```{r}
heatmap_all_all <- results_heatmap %>% filter(`Data Subset` == "All") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  labs(
    title = "All Predictors",
    x = "Top Features",
    y = "Model Configuration"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),
    legend.position = "none"
  )

heatmap_all_inf <- results_heatmap_infra %>% filter(`Data Subset` == "All") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Infrastructure",
    x = "",
    y = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank()
  )

heatmap_all_visual <- heatmap_all_all + heatmap_all_inf + plot_layout(ncol = 2, widths = c(2, 1))
```

```{r}
#| label: figure-13a
#| fig-cap: "**Figure 13a:** Feature importance heatmap across model configurations (all crashes)"
#| fig-align: center
heatmap_all_visual
```

When analyzing models trained on the complete crash dataset, distinct patterns emerge in the predictors for different severity targets. Notably, the Fatal vs. Non-Fatal classification models consistently identify Drug Involvement as the most influential predictor across all model types. This variable is followed in importance by Crash Speed Involvement and Alcohol Involvement, suggesting that these behavioral factors are key determinants of fatal crash outcomes in the full dataset.

Models targeting Degree of Injury outcomes exhibit similar preferences, frequently selecting Drug Involvement and Alcohol Involvement as top-ranked features. However, both Degree of Injury and Injured vs Non-Injured models also rank road condition features such as “Road Surface – Unknown” highly, particularly in LASSO and Ridge regression. This repeated prominence suggests a potential association between uncertain or missing surface condition data and injury severity, possibly reflecting either a true risk factor or a proxy for incomplete contextual information.

In contrast to the linear models, Random Forests tended to emphasize numeric and continuous features such as Total Watts, Speed Limit, and Crash Hour. This divergence in feature selection likely reflects the Random Forest’s capacity to capture non-linear relationships and complex interactions that are less accessible to linear or penalized regression approaches.

In the models that were trained on infrastructural features only, the presence of other infrastructure features (Infrastructure Present) emerges as particularly important in predicting injury outcomes. For models predicting Fatal vs. Non-Fatal outcomes using infrastructure-only features, the Sum of Traffic Control Devices appears as the most predictive feature, suggesting a potential mitigating effect of traffic-calming infrastructure on crash fatality.


## Modeling Results - Bicycle Only Crashes
```{r}
heatmap_bike_all <- results_heatmap %>% filter(`Data Subset` == "Bike") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "All Predictors",
    x = "Top Features",
    y = "Model Configuration"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f4f8ff"),
    legend.background = element_rect(fill = "#f4f8ff")
  )

heatmap_bike_inf <- results_heatmap_infra %>% filter(`Data Subset` == "Bike") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Infrastructure",
    x = "",
    y = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "#f4f8ff"),
    legend.background = element_rect(fill = "#f4f8ff")
  )

heatmap_bike_visual <- heatmap_bike_all + heatmap_bike_inf + plot_layout(ncol = 2, widths = c(2.25, 0.75))
```

```{r}
#| label: figure-13b
#| fig-cap: "**Figure 13b:** Feature importance heatmap across model configurations (bicycle crashes)"
#| fig-align: center
heatmap_bike_visual
```

Models trained exclusively on bicycle-involved crashes exhibit patterns largely consistent with those observed in the full crash dataset. In particular, for the Fatal vs. Non-Fatal outcome, Drug Involvement remains the most influential predictor across all model configurations. This variable also appears frequently among top-ranked features in models predicting Degree of Injury, alongside Alcohol Involvement and Crash Speed Involvement, reinforcing the significance of behavioral factors in severe bicycle-related crash outcomes.

Within the Degree of Injury models, both LASSO and Ridge regressions highlight snowy or icy road conditions as among the most predictive features. This finding suggests that adverse environmental conditions may play a heightened role in determining injury severity when bicycles are involved, potentially due to cyclists’ increased vulnerability to surface hazards.

Overall, feature rankings in the bicycle-only models remain broadly consistent with those trained on the full dataset, particularly in the prioritization of behavioral and environmental risk factors.

In contrast, the infrastructure-only models for bicycle crashes do not reveal strong or consistent patterns across target outcomes. Nevertheless, some model-specific tendencies emerge: Bike Lane presence is frequently identified as a top feature in linear models, while Total Watts appears as a dominant predictor in Random Forest models. These results suggest that while infrastructure variables may influence crash severity for cyclists, their effects may be more context-dependent and less universally predictive than behavioral or environmental factors.

## Modeling Results - Pedestrian Only Crashes
```{r}
heatmap_ped_all <- results_heatmap %>% filter(`Data Subset` == "Ped") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "All Predictors",
    x = "Top Features",
    y = "Model Configuration"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f4f8ff"),
    legend.background = element_rect(fill = "#f4f8ff")
  )

heatmap_ped_inf <- results_heatmap_infra %>% filter(`Data Subset` == "Ped") %>%
ggplot(aes(x = Feature, y = fct_rev(Model_ID), fill = as.factor(Importance_Rank))) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c("1" = "#1b9e77", "2" = "#d95f13", "3" = "#7570b3"),
    name = "Rank"
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Infrastructure",
    x = "",
    y = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "#f4f8ff"),
    legend.background = element_rect(fill = "#f4f8ff")
  )

heatmap_ped_visual <- heatmap_ped_all + heatmap_ped_inf + plot_layout(ncol = 2, widths = c(2.1, 0.9))
```

```{r}
#| label: figure-13c
#| fig-cap: "**Figure 13c:** Feature importance heatmap across model configurations (pedestrian crashes)"
#| fig-align: center
heatmap_ped_visual
```

Pedestrian-involved crash models continue to reflect patterns observed in the broader dataset, with behavioral features playing a central role in predicting crash severity. Across models targeting both Fatal vs. Non-Fatal and Degree of Injury outcomes, Drug Involvement remains the most frequently selected feature. Notably, in regularized models (LASSO and Ridge), Marijuana Involved consistently ranks as the most influential predictor, suggesting a potential substance-specific association with pedestrian crash outcomes.

While no dominant pattern emerges in the Injured vs. Non-Injured models, certain features such as Driveway or Alley presence and the Sum of Traffic Calming Devices appear with some regularity in the linear models. Additionally, for Fatal vs. Non-Fatal predictions, the Hit and Run indicator is frequently identified as a top feature, reflecting its potential link to crash severity in pedestrian incidents.

In the infrastructure-only models, Bike Lane presence emerges as a leading predictor in Degree of Injury models using logistic regression, with broader infrastructure presence indicators following closely. In the Injured vs. Non-Injured models, the Sum of Traffic Calming Devices appears again as a contributing feature. For Fatal vs. Non-Fatal predictions, Speed Limit is the most significant predictor in logistic regression models and ranks second in Random Forest models. Total Watts also appears consistently across all three target outcomes, underscoring the potential role of visibility and lighting in pedestrian crash severity.

## Summary of Results
Taken together, the results across all crash types and modeling configurations reveal several consistent insights into the factors most associated with crash severity. Behavioral factors, particularly drug involvement, emerge as the most influential predictors of fatal outcomes across all road user types. This pattern is especially pronounced in regularized models, where alcohol involvement and crash speed consistently appear as a top predictor.

Across degree of injury and injury vs. non-injury outcomes, the influence of environmental and contextual features becomes more apparent. Variables such as road surface conditions (especially when unknown or icy), lighting conditions (Total Watts), and crash timing (Crash Hour) rank highly, particularly in Random Forest models capable of capturing non-linear interactions. Notably, road surface unknown appears prominently in linear models, raising the possibility that data gaps or ambiguous reporting may be correlated with crash severity.

Infrastructure-only models, while generally less predictive than those using the full feature set, still highlight important built environment characteristics. Features such as the presence of bike lanes, traffic calming devices, and speed limits frequently emerge as top predictors of injury outcomes. For fatality predictions, traffic control infrastructure appears to play a meaningful role, suggesting that strategically placed interventions may help mitigate the most severe outcomes. The variation in feature importance across model types further underscores the value of using multiple modeling approaches to gain complementary perspectives on risk factors.

Overall, the findings support an understanding of crash severity where behavioral interventions, environmental awareness, and infrastructure improvements all have roles to play. By disaggregating analyses by crash type and outcome severity, these results offer a more granular view of the conditions under which crashes result in serious harm. These insights can directly inform targeted safety strategies for motorists, bicyclists, and pedestrians alike.

